---
title: "p8105_hw1_jt3466"
author: "Johnstone Tcheou"
date: "2024-09-12"
output: github_document
---

# Homework 1

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
data("penguins", package="palmerpenguins")
```

## Problem 1

The penguins dataset has already been installed and loaded, named penguins.

```{r summary}
summary(penguins)
```
The penguins dataset has `r nrow(penguins)` rows and `r ncol(penguins)`, consisting of data on penguins like species, island, body measurements, sex, and year. The mean flipper length is `r mean(penguins$flipper_length_mm, na.rm=TRUE)`.

```{r scatterplot, warning=FALSE}
ggplot(penguins, aes(x=bill_length_mm, y=flipper_length_mm, color=species)) + 
  geom_point()

ggsave("scatterplot.pdf")
```

This is a scatterplot of flipper length in mm against bill length in mm, with colors corresponding to the species of penguin. There are 2 observations with NA for bill and flipper lengths, so the warning that these observations were removed have been suppressed in this code chunk.

## Problem 2

```{r dataframe}
test_df <- tibble(
  rsamp=rnorm(10),
  vec_logical=rsamp>0,
  vec_char=c("This", "vector", "is", "a", "character", "vector", "with", "a", "length", "10"),
  vec_factor=factor(c("easy","medium","hard","medium","medium","easy","hard","hard","easy","medium"))
)

mean(test_df$rsamp) 
mean(test_df$vec_logical) 
mean(test_df$vec_char)
mean(test_df$vec_factor) 
```

Taking the means of the random sample and the logical vector works.The random sample is of all numeric values, and the mean of the logical vector works because TRUE is numerically 1 and FALSE is numerically 0, so being a random normally distributed samples, these average out to 0.5.

However, the means of the character and factor vectors do not work, as these cannot be internally be coerced to numeric. 

```{r coercion}

test_df |>
  mutate(
    rsamp=as.numeric(rsamp),
    vec_logical=as.numeric(vec_logical),
    vec_char=as.numeric(vec_char),
    vec_factor=as.numeric(vec_factor)
  )
```

The output retains the random sampling values, coerces the boolean values to corresponding 1s and 0s, and cannot coerce character or factor values to numeric. NAs are introduced instead, showing why the means cannot be calculated for these variables that cannot be coerced to numeric variable type.